Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}

### 2017-08-01 revision/review_response.Rnw

\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}

\begin{document}
\raggedright

% \title{}
% \author{}
% \maketitle

{\it Response to reviewer comments ``Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern Gulf of Mexico'', Beck MW, Lehrter JC, Lowe LL, Jarvis BM.}

{\it We thank the editor and reviewer for providing thoughtful comments on our manuscript.  Responses to these comments are shown in italics.  Page and paragraph numbers refer to the revised manuscript.}

@@

\Bigtxt{Editor's comments:}

The authors executed the work on parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern Gulf of Mexico. The authors did not provide proper conceptual diagram of the model.

{\it }

The model is not presented in detail, it is necessary to give more details on original model.

{\it }

Discussion part is unnecessary long, lot of introduction in the discussion. This part needs major revisions and complete restructure.

{\it The discussion was shortened and introductory material was removed.  Specifically, the first paragraph of the discusison was removed and the first paragraph of sec. 4.1 was shortened.  This reduced the length by over a page.}

The sensitivity analysis was adequately implemented but further clarification is needed for how the water quality model was applied in Weeks Bay.

{\it }

In view of the above comments and reports of reviewers, your manuscript has been evaluated and you are informed to resubmit the manuscript after proper and thorough revisions in accordance with the comments of editor/reviewers. Review reports are appended below. Thank you for your submission in Ecological Modelling.

@@

For the sensitivity analysis the authors manipulated the observations of DO in the estuary filtering the tidal impacts to be able to implement the FishTank model. However, there is no evidence of the resulting time series of DO or of the other variables analyzed. I am concerned that while removing the DO variations due to tidal cycles, the authors could have also removed the daily variation of DO which can be associated to diurnal cycles of sunlight and temperature. If this is the case, then the DO calibration is based almost on synthetic and unrealistic DO data. The authors must include a section showing the data that was used for the calibration analysis. This section should include a location figure showing Weeks Bay, AL, the stations used to extract the original datasets, plots of the unprocessed and processed time series of DO, ammonium, nitrate, chl-a and irradiance as well as plots showing the processed data against the calibrated model outputs for each variable. The authors must also include a table with the calibrated parameter values. This information is standard in all modeling studies.

Please also describe what are the forcing conditions of the FishTank model. There are two tributary rivers in Weeks Bay and one outlet condition. How is each one simulated in your FishTank model?.

Other comments:

Introduction. I suggest the authors to include also references of couple hydrodynamic and water quality studies available in the region. For example, the work of Camacho et al. (2014a) in St. Louis Bay, MS and part of Mississippi Sound is a relevant citation that is missing in this investigation. The work of Dortch et al. (2007) in Mississippi Sound should also be cited.

{\it Citations for Dortch et al. 2007 and Camacho et al. 2014a were added to the first paragraph.}

Sensitivity analyses have also been conducted in the past in the region. For example, Camacho et al (2014b) conducted an uncertainty and sensitivity analysis of a hydrodynamic model of Weeks Bay Alabama. I suggest the authors to include these references. The sensitivity estimates provided by Camacho et al. (2014b) seem to be based on the same equation Eq. (1) included in this manuscript. Can the authors briefly indicate (perhaps one or two sentences) if both methods are the same?

{\it Yes, this approach is similarly described by Camacho et al. 2014b.  The following text was added to paragraph one in sec. 2.2: `This approach is described as a First Order Variance Analysis that linearly propagates changes from model parameters to model predictions (Camacho et al. 2014b).'}

@@

lines 24 - 39 page 3. This part of the text is full of vague statements. Please be more explicit. What are the ``characteristics expected from the output'' and the ``information represented by the structural components''. What do you mean with ``Given that these characteristics cannot be simultaneously achieved, models are developed in partial dependence of reality and theoretical constructs, completely separate from both, or dependent on one or the other''.

{\it This section was revised as follows:

`The development of a model represents a tradeoff between achieving predictive accuracy and a realistic representation of environmental processes.... Given that these characteristics cannot be simultaneously achieved, models are developed that balance predictive accuracy with environmental realism, often favoring one at the expense of the other.'

}

@@

lines 24 - 39 page 3. I think the reference Levins (1966) is extremely old and must be removed.

{\it This citation was referenced in Ganju et al. 2016, from which this information was obtained.  We have not removed the citation because we feel that age does not make a publication inadequate.  However, we have added the citation to Ganju et al. 2016 as a more current reference.}

@@

It is interesting to see that the authors constantly use the term ``precision'' to denote goodness of fit. In general, precision is used to denote the number of digits of a number. Note that a number can be precise but inaccurate. I suggest the authors to revise the text and replace ``precision'' with a better term to denote goodness of fit. e.g. level of agreement between observations and simulations.

{\it We have found that precision is commonly used in the literature to describe goodness of fit, although we appreciate the distinction made by the reviewer.  Considering an example where observed data represent a biased sample of an environmental process (e.g., infrequent or unevent sampling of an event that varies over time), model calibration to these data can only possibly achieve precision.  Although practitioners would not knowingly use observed data that are biased for model calibration, we are certain that many ecological datasets are incomplete representations of actual processes.  We argue then that model calibration in many cases is better described as seeking precision, although in practice it is stated as predictive accuracy given an incomplete knowledge of the observed data relative to the true environmental process. Regardless, we have replaced precision with ``goodness of fit'' or ``fit'' as a more generic term.}

@@

Line 9 page 10. Selection or Selected?

{\it Changed to: `Heurestics for parameter selection also recognized...'}

@@

Line 4 page 12. what are the default parameter conditions?. You should include a table with the default parameter values and parameter values under 50\% modified conditions.

Lines 29-30 page 16. Please include the units in the reported RMSEs. mmol O2?

{\it RMSE units are reported as mmol o2 m-3 and was added to the text.}

### 2017-08-01 ident_manu.Rnw
revisions half way done

# library(Jabbrev)
# bbl_ext('ident_manu.bbl', 'refs.bib', 'ident_manu.bib')
# bib_shr('ident_manu.bib', 'refsshr.bib')
@

@@

\begin{abstract}
\noindent Local sensitivity analyses and identifiable parameter subsets were used to describe numerical constraints of a hypoxia model for bottom waters of the northern Gulf of Mexico.  The sensitivity of state variables differed considerably with parameter changes, although most variables were responsive to changes in parameters that influenced planktonic growth rates and less sensitive to physical or chemical parameters.  Variation in sensitivity had a direct correspondence with identifiability, such that only small subsets of the complete parameter set had unique effects on the model output. Selecting parameters by decreasing sensitivity demonstrated that only eight of \Sexpr{sum(unlist(nparms))} total parameters had a sufficiently unique effect on model output for precise calibration.  As a result, parameter selection heuristics were used to identify parameters for model calibration that depended on combined effects on output, relative sensitivity of each parameter, and ecological categories for the biogeochemical equations. The calibrated \ac{zerod} unit of the hypoxia model had improved precision if sensitive parameters were included in an identifiable subset.  Extension of results to a three-dimensional grid of the Gulf of Mexico showed that sensitive parameters for the \ac{zerod} model translated to non-trivial changes in the areal estimates of hypoxia.
\end{abstract}

@@

Hypoxia formation in bottom waters of coastal oceans occurs primarily from excess nutrient inputs from land-based sources \citep{Justic87,Diaz95,Howarth96}.  These events are detrimental to aquatic organisms and have significant negative effects on economic resources derived from coastal ecosystems \citep{Lipton03,Diaz11}.  An understanding of the biological, physical, and chemical processes that influence hypoxic areas is a critical concern for mitigating and preventing these negative impacts.  Numerical ecosystem models are important tools that synthesize knowledge of ecosystem processes that contribute to hypoxia formation and for predicting the effects of proposed management activities or future scenarios \citep{Scavia04,Hagy07,Pauer16}.  Unlike statistical models with more generic structures, simulation and process-based models include explicit descriptions of relevant processes that are constrained by empirical or observational data relevant to the system of interest \citep[e.g.][]{Omlin01b,Eldridge10}.  These models are often coupled with hydrodynamic grids to provide spatially-explicit representations of patterns in three dimensions \citep{Warner05,Zhao10,Ganju16}. Combined hydrodynamic and bio-geo-chemical models have been developed specifically to describe hypoxic conditions on the \ac{lcs} in the northern \ac{gom} (\citealt{Fennel13,Obenour15,Pauer16}, \citetalias{LehrterIP}).  This area drains a significant portion of the continental United States through the \ac{marb} and is the second largest hypoxic area in the world \citep{Rabalais02}.  Understanding processes that contribute to the frequency and duration of hypoxic events remains a critical research goal for the region.

The development of a model represents a tradeoff between characteristics expected from the output and information represented by the structural components. An ideal model is sufficiently generalizable across systems, provides results that are precise given the inputs, and includes components that are realistic descriptions of actual processes \citep{Levins66}. Given that these characteristics cannot be simultaneously achieved, models are developed in partial dependence of reality and theoretical constructs, completely separate from both, or dependent on one or the other \citep{Morrison99,Ganju16}.  These challenges are analagous to the well-known bias-variance tradeoff in statistical models that balances the competing objectives of over- and under-fitting to an observed dataset. Process-based models are more commonly imbalanced between reality and theory, such that most are over-parameterized in an attempt to completely describe reality \citep{Denman03,Nossent12,Petrucci14}.  Quantitative limitations of over-parameterization are analagous to degrees of freedom in standard statistical models as free parameters cannot be numerically estimated when constrained to an observed dataset \citep{Kirchner06}.  More importantly, over-parameterization can limit use across systems outside of the data domain and impose uncertainty in model predictions as realistic values for every variable may not be known or inaccurately applied from existing studies \citep{Durand02,Refsgaard07,Wade08}.

Model precision can be evaluated relative to the effects of initial conditions or the observed data used for calibration, changes in parameter values, or variation in the structural components (i.e., observational, parameter, or structural uncertainty) \citep{Beck87}.  Evaluating effects of parameter changes is by far the most common and simplest approach.  Although sensitivity analyses should be integrated with model development, parameters are often evaluated post-hoc as a form of `damage control' for further calibration.  This approach is sometimes called inverse modelling where results from sensitivity analyses are used to guide calibration or fit of the developed model to observations (\citealt{Soetaert10}, or confronting models with data, \textit{sensu} \citealt{Hilborn97}).  Parameter sensitivity analysis combined with inverse modelling necessarily involves questions of parameter `identifiability'.  Redundancies in parameter effects lead to unidentifiable models where calibration is empirically impossible (i.e., standard algorithms will not converge) or parameter values may be non-unique leading to the right answer for the wrong reason \citep{Kirchner06}. Unidentifiable parameter sets have effects on model output that can be undone or compensated for by alteration of other parameters.  Identifiability issues are not foreign to hypoxia or eutrophication models  \citep{Omlin01,Estrada10,Mateus15}, although there is a clear need for greater integration of these concepts in practice \citep{Fasham06}.

This study describes a sensitivity and identifiability analysis of a \ac{zerod} unit of a larger spatial-temporal model of hypoxia dynamics on the \ac{lcs}.  The objectives were to provide a statistical approach that demonstrates numerical limitations of parameter sets for model calibration and provide a framework for selecting parameters within the identifiability constraints.  The specific goals were to \begin{inparaenum}[1\upshape)]
\item identify the parameters that have the greatest influence on state variables using local sensitivity analysis,
\item quantify the identifiability of subsets of the total parameter space based on sensitivity,
\item and provide a set of heuristics for choosing parameters based on sensitivity, identifiability, and parameter categories.
\end{inparaenum}
The \ac{zerod} model was calibrated with selected parameter subsets to demonstrate use of the selection heuristics to improve model precision.  Sensitive parameters for the \ac{zerod} model were also varied in the larger 3-dimensional model to demonstrate scalability the results. In addition to \ac{do}, other state variables that were evaluated included ammonium, \ac{chla}, irradiance, nitrate, \ac{pom}, \ac{dom}, and phosphorus. In general, we provide empirical results to support the assumption that models are generally over-parameterized and only a finite and smaller subset of the complete parameter set can be optimized.

@@

A local sensitivity analysis was performed by evaluating the change in state variables following perturbation of each parameter from its original value \citep{Soetaert10,RDCT17}. Parameters were individually perturbed by a 50\% increase of the original values and sensitivity $S$ was estimated for each time step $i$ given a change for parameter $j$ as:

@@

\noindent where $\gamma$ ranges from one to infinity for perfectly identifiable (orthogonal) or unidentifiable (perfectly collinear) parameter sets.  The sensitivity functions were supplied as a matrix $\hat{S}$ with rows $i$ and columns $j$ (\cref{sijeqn}) that described deviations of predicted \ac{do} following perturbations of each parameter.  Thus, $\gamma$ can be estimated from results for any subset of parameter combinations. Sensitivity matrices were first normalized by dividing by the square root of the summed residuals \citep{Omlin01,Soetaert10}. Estimates of $\gamma$ greater than 10-15 suggest parameter sets are poorly identifiable \citep{Brun01,Omlin01}, meaning parameter values that maximize precision on a calibration dataset are inestimable by conventional optimization algorithms. An intuitive interpretation of $\gamma$ is provided by \citet{Brun01}, such that a change in a state variable caused by a change in one parameter can be offset by the fraction $1 - 1/\gamma$ by the remaining parameters.  That is, $\gamma = 10$ suggests the relative change in \ac{do} for an arbitrary parameter in the selected set can be compensated for by 90\% with changes in the other parameters.

Parameter selection for model calibration must consider the competing objectives of increased precision with parameter inclusion and reduced identifiability as it relates to optimization.  An additional challenge is a large number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter (e.g., parameters for a group of related structural equations could be of interest).  \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., $n$ choose $k$ combinations, $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.

A set of heuristics was developed that addresses the tradeoff in model complexity and identifiability given the challenges described above \citep[see also][]{Wagener01}.  These rulesets were developed to select parameters with preference for those with high sensitivity and identifiability based on $\gamma < 15$ as an acceptable threshold for subsets (e.g., 93\% accountability between parameters).  Selection heurestics also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences by users given desired application of the model.  For all selection heuristics, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item blocked within each of five parameter categories,
\item independent of parameter category,
\item or considering all categories equally.
\end{inparaenum}

@@

Selected parameter subsets from the seven heuristics were used to calibrate the \ac{zerod} model.  This analysis demonstrated that different parameter sets can produce model output with differing precision and calibration effort given
\begin{inparaenum}[1\upshape)]
\item differences in variable sensitivity to parameters in each subset, 
\item differences in the parameters selected from different categories (heuristic difference), and 
\item differences in identifiability values.
\end{inparaenum}
Generally, this analysis was a proof of concept that the selection heuristics can lead to an `optimal' model by maximizing the balance between sensitivity and identifiability.

@@

Each parameter subset was calibrated to most closely match the model output to the observed time series.  Parameter values were searched using the optim function in R that implemented the limited memory modification of the BFGS quasi-Newton method \citep{Byrd95,Nocedal06,RDCT17}.  This optimization scheme searches for values within the user-defined lower and upper bounds of each parameter.  We set the algorithm to search within +/- 50\% of the default parameter values.  While we recognize that `actual' parameter values could possibly extend beyond this range, constraining the values to the range tested herein was expected to produce output that was more interpretable within the context of the sensitivity functions defined above.  Moreover, precise limits on each parameter were less of a concern for the current analysis as our primary goal was to demonstrate differences in precision as a function of sensitivity and identifiability.  The optimal values for each parameter were identified based on convergence of the model results as a minimization of the \ac{rmse} between the observed and modelled data.

@@

Calibration of parameter values for each of the seven selection heuristics (\cref{tab:heurist1,tab:heurist2,tab:heurist3})  produced model output that varied in precision and computational effort required to identify optimal values (\cref{tab:calib}).  As expected, the number of required iterations of the optimization algorithm to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only \Sexpr{toeng(temp$its)[[1]]} iterations were required to optimize the single parameter selected for the temperature category, whereas \Sexpr{top$its} iterations were required for the \Sexpr{toeng(top$nprm)[[1]]} parameters selected independent of category.  No association was observed between collinearity ($\gamma$) and number of iterations or calibration.  Of the seven parameter subsets, selection independent of parameter category and equally within each category produced the largest reduction in \ac{rmse} values from the default output (starting \ac{rmse} \Sexpr{strt}, reduction to \Sexpr{scinot(top$rmse)}, \Sexpr{scinot(cattop$rmse)}, respectively), whereas the optimal values for parameters in the Optics category had the smallest reduction (\Sexpr{scinot(optics$rmse)}).  Lower \ac{rmse} was associated with higher average sensitivity of a parameter subset.  Parameters selected only in the optics category had the highest \ac{rmse} and lowest average L1 (\Sexpr{scinot(optics$avee)}, \Sexpr{toeng(optics$nprm)} parameters), whereas parameters selected independent of category had the lowest \ac{rmse} and highest average L1 (\Sexpr{scinot(top$avee)}, \Sexpr{toeng(top$nprm)} parameters, \cref{tab:calib}).

@@

Common goals in the application of biogeochemical models are to \begin{inparaenum}[1\upshape)]
\item accurately describe the system by matching predictions with observed data \citep{Reckhow90}, and 
\item provide a means of forecasting ecosystem condition with hypothesized management or environmental scenarios \citep{Clark01}. 
\end{inparaenum}
Although these objectives are the focus of most applications, the structural components and associated parameters of process-based models should secondarily provide insight of ecosystem processes that are driving observed changes.  This latter objective represents a more generic scientific purpose of modelling that extends beyond the benefits of describing and predicting change.  Modelers aim to identify universal principles that govern dynamics across systems and the constraint of model parameters to observations provides a means of supporting or refuting hypotheses \citep{Hilborn97,Kirchner06}.  Accordingly, this study provided a simple approach to use the effects of parameter perturbations on state variables to characterize identifiable parameter subsets that varied by selection criteria.  By doing so, we provided a simple and preliminary approach towards improved model refinement that begins to address more general purposes of model applications. Our specific results demonstrated that small parameter subsets relative to all sensitive parameters were within the identifiability thresholds described in the literature.  The identifiable parameter subsets varied considerably between state variables and the method for parameter selection.  Use of identifiable parameter subsets with sensitive parameters improved model precision and results for the smaller model were scalable to the 3-dimensional model.  In general, these results provide justification for the use of explicit parameter selection heuristics to improve model precision.

State variables were most sensitive to parameters in the phytoplankton and zooplankton categories, particularly the maximum growth rates (\Sexpr{par_txt('umax_1')} for phytoplankton, \Sexpr{par_txt('Zumax_1')} for zooplankton), mortality coefficient for phytoplankton (\Sexpr{par_txt('mA_1')}), and the zooplankton half saturation coefficent for grazing (\Sexpr{par_txt('ZKa_1')}). An increase in the growth rate of primary producers has the potential to increase oxygen concentration through photosynthetic processes, although increased production of organic matter is balanced with respiration and bacterial decomposition that reduce \ac{do} in the water column.  Similarly, increases in zooplankton abundance with increased growth rates causes a reduction in phytoplankton biomass through grazing, which is expected to further deplete pools of organic matter from algal sources.  Most variables were also sensitive to variation in the half-saturation grazing coefficient which moderates the concentration of nutrients that support half the maximum grazing rate. Although the tradeoff between abundance, grazing, and decomposition is complex, the sensitivity of model state variables to parameters that directly control the abundance of primary producers is in agreement with empirical observations of factors that influence hypoxia dynamics on the \ac{lcs} \citep{Fahnenstiel95,Roelke00,Eldridge10}. The sensitivity of the model output to variation in other parameters that relate to physical and chemical properties of the system was of secondary importance to biological relationships.  That is, state variables were sensitive to changes in light and temperature parameters, although to a lesser extent than phytoplankton and zooplankton parameters.  As such, the differing sensitivities of state variables to parameters in each of the categories was not unexpected given general ecological relationships that are well understood and described by the model.

>>>>

# # library(Jabbrev)
# bbl_ext('ident_manu.bbl', 'refs.bib', 'ident_manu.bib')
# bib_shr('ident_manu.bib', 'refsshr.bib')
@

@@

\begin{abstract}
\noindent Local sensitivity analyses and identifiable parameter subsets were used to describe numerical constraints of a hypoxia model for bottom waters of the northern Gulf of Mexico.  The sensitivity of state variables differed considerably with parameter changes, although most variables were responsive to changes in parameters that influenced planktonic growth rates and less sensitive to physical or chemical parameters.  Variation in sensitivity had a direct correspondence with identifiability, such that only small subsets of the complete parameter set had unique effects on the model output. Selecting parameters by decreasing sensitivity demonstrated that only eight of \Sexpr{sum(unlist(nparms))} total parameters had a sufficiently unique effect on model output for accurate calibration.  As a result, parameter selection heuristics were used to identify parameters for model calibration that depended on combined effects on output, relative sensitivity of each parameter, and ecological categories for the biogeochemical equations. The calibrated \ac{zerod} unit of the hypoxia model had improved fit to the observed data if sensitive parameters were included in an identifiable subset.  Extension of results to a three-dimensional grid of the Gulf of Mexico showed that sensitive parameters for the \ac{zerod} model translated to non-trivial changes in the areal estimates of hypoxia.
\end{abstract}

@@

Hypoxia formation in bottom waters of coastal oceans occurs primarily from excess nutrient inputs from land-based sources \citep{Justic87,Diaz95,Howarth96}.  These events are detrimental to aquatic organisms and have significant negative effects on economic resources derived from coastal ecosystems \citep{Lipton03,Diaz11}.  An understanding of the biological, physical, and chemical processes that influence hypoxic areas is a critical concern for mitigating and preventing these negative impacts.  Numerical ecosystem models are important tools that synthesize knowledge of ecosystem processes that contribute to hypoxia formation and for predicting the effects of proposed management activities or future scenarios \citep{Scavia04,Hagy07,Camacho14a,Pauer16}.  Unlike statistical models with more generic structures, simulation and process-based models include explicit descriptions of relevant processes that are constrained by empirical or observational data relevant to the system of interest \citep[e.g.][]{Omlin01b,Eldridge10}.  These models are often coupled with hydrodynamic grids to provide spatially-explicit representations of patterns in three dimensions \citep{Warner05,Dortch07,Zhao10,Ganju16}. Combined hydrodynamic and bio-geo-chemical models have been developed specifically to describe hypoxic conditions on the \ac{lcs} in the northern \ac{gom} (\citealt{Fennel13,Obenour15,Pauer16}, \citetalias{LehrterIP}).  This area drains a significant portion of the continental United States through the \ac{marb} and is the second largest hypoxic area in the world \citep{Rabalais02}.  Understanding processes that contribute to the frequency and duration of hypoxic events remains a critical research goal for the region.

The development of a model represents a tradeoff between achieving predictive accuracy and a realistic representation of environmental processes. An ideal model is sufficiently generalizable across systems, provides results that are accurate given the inputs, and includes components that are valid descriptions of actual processes \citep{Levins66, Ganju16}. Given that these characteristics cannot be simultaneously achieved, models are developed that balance predictive accuracy with environmental realism, often favoring one at the expense of the other \citep{Morrison99,Ganju16}.  These challenges are analagous to the well-known bias-variance tradeoff in statistical models that balances the competing objectives of over- and under-fitting to an observed dataset. Process-based models are more commonly imbalanced between reality and theory, such that most are over-parameterized in an attempt to completely describe reality \citep{Denman03,Nossent12,Petrucci14}.  Quantitative limitations of over-parameterization are analagous to degrees of freedom in standard statistical models as free parameters cannot be numerically estimated when constrained to an observed dataset \citep{Kirchner06}.  More importantly, over-parameterization can limit use across systems outside of the data domain and impose uncertainty in model predictions as realistic values for every variable may not be known or inaccurately applied from existing studies \citep{Durand02,Refsgaard07,Wade08}.

Model fit can be evaluated relative to the effects of initial conditions or the observed data used for calibration, changes in parameter values, or variation in the structural components (i.e., observational, parameter, or structural uncertainty) \citep{Beck87}.  Evaluating effects of parameter changes is by far the most common and simplest approach.  Although sensitivity analyses should be integrated with model development, parameters are often evaluated post-hoc as a form of `damage control' for further calibration.  This approach is sometimes called inverse modelling where results from sensitivity analyses are used to guide calibration or fit of the developed model to observations (\citealt{Soetaert10}, or confronting models with data, \textit{sensu} \citealt{Hilborn97}).  Parameter sensitivity analysis combined with inverse modelling necessarily involves questions of parameter `identifiability'.  Redundancies in parameter effects lead to unidentifiable models where calibration is empirically impossible (i.e., standard algorithms will not converge) or parameter values may be non-unique leading to the right answer for the wrong reason \citep{Kirchner06}. Unidentifiable parameter sets have effects on model output that can be undone or compensated for by alteration of other parameters.  Identifiability issues are not foreign to hypoxia or eutrophication models  \citep{Omlin01,Estrada10,Mateus15}, although there is a clear need for greater integration of these concepts in practice \citep{Fasham06}.

This study describes a sensitivity and identifiability analysis of a \ac{zerod} unit of a larger spatial-temporal model of hypoxia dynamics on the \ac{lcs}.  The objectives were to provide a statistical approach that demonstrates numerical limitations of parameter sets for model calibration and provide a framework for selecting parameters within the identifiability constraints.  The specific goals were to \begin{inparaenum}[1\upshape)]
\item identify the parameters that have the greatest influence on state variables using local sensitivity analysis,
\item quantify the identifiability of subsets of the total parameter space based on sensitivity,
\item and provide a set of heuristics for choosing parameters based on sensitivity, identifiability, and parameter categories.
\end{inparaenum}
The \ac{zerod} model was calibrated with selected parameter subsets to demonstrate use of the selection heuristics to improve model fit.  Sensitive parameters for the \ac{zerod} model were also varied in the larger 3-dimensional model to demonstrate scalability the results. In addition to \ac{do}, other state variables that were evaluated included ammonium, \ac{chla}, irradiance, nitrate, \ac{pom}, \ac{dom}, and phosphorus. In general, we provide empirical results to support the assumption that models are generally over-parameterized and only a finite and smaller subset of the complete parameter set can be optimized.

@@

A local sensitivity analysis was performed by evaluating the change in state variables following perturbation of each parameter from its original value \citep{Soetaert10,Camacho14b,RDCT17}. This approach is described as a First Order Variance Analysis that linearly propagates changes from model parameters to model predictions \citep{Camacho14b}. Parameters were individually perturbed by a 50\% increase of the original values and sensitivity $S$ was estimated for each time step $i$ given a change for parameter $j$ as:

@@

\noindent where $\gamma$ ranges from one to infinity for perfectly identifiable (orthogonal) or unidentifiable (perfectly collinear) parameter sets.  The sensitivity functions were supplied as a matrix $\hat{S}$ with rows $i$ and columns $j$ (\cref{sijeqn}) that described deviations of predicted \ac{do} following perturbations of each parameter.  Thus, $\gamma$ can be estimated from results for any subset of parameter combinations. Sensitivity matrices were first normalized by dividing by the square root of the summed residuals \citep{Omlin01,Soetaert10}. Estimates of $\gamma$ greater than 10-15 suggest parameter sets are poorly identifiable \citep{Brun01,Omlin01}, meaning parameter values that maximize fit on a calibration dataset are inestimable by conventional optimization algorithms. An intuitive interpretation of $\gamma$ is provided by \citet{Brun01}, such that a change in a state variable caused by a change in one parameter can be offset by the fraction $1 - 1/\gamma$ by the remaining parameters.  That is, $\gamma = 10$ suggests the relative change in \ac{do} for an arbitrary parameter in the selected set can be compensated for by 90\% with changes in the other parameters.

Parameter selection for model calibration must consider the competing objectives of increased goodness of fit with parameter inclusion and reduced identifiability as it relates to optimization.  An additional challenge is a large number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter (e.g., parameters for a group of related structural equations could be of interest).  \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., $n$ choose $k$ combinations, $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.

A set of heuristics was developed that addresses the tradeoff in model complexity and identifiability given the challenges described above \citep[see also][]{Wagener01}.  These rulesets were developed to select parameters with preference for those with high sensitivity and identifiability based on $\gamma < 15$ as an acceptable threshold for subsets (e.g., 93\% accountability between parameters).  Heurestics for parameter selection also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences by users given desired application of the model.  For all selection heuristics, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item blocked within each of five parameter categories,
\item independent of parameter category,
\item or considering all categories equally.
\end{inparaenum}

@@

Selected parameter subsets from the seven heuristics were used to calibrate the \ac{zerod} model.  This analysis demonstrated that different parameter sets can produce model output with differing goodness and fit and calibration effort given
\begin{inparaenum}[1\upshape)]
\item differences in variable sensitivity to parameters in each subset, 
\item differences in the parameters selected from different categories (heuristic difference), and 
\item differences in identifiability values.
\end{inparaenum}
Generally, this analysis was a proof of concept that the selection heuristics can lead to an `optimal' model by maximizing the balance between sensitivity and identifiability.

@@

Each parameter subset was calibrated to most closely match the model output to the observed time series.  Parameter values were searched using the optim function in R that implemented the limited memory modification of the BFGS quasi-Newton method \citep{Byrd95,Nocedal06,RDCT17}.  This optimization scheme searches for values within the user-defined lower and upper bounds of each parameter.  We set the algorithm to search within +/- 50\% of the default parameter values.  While we recognize that `actual' parameter values could possibly extend beyond this range, constraining the values to the range tested herein was expected to produce output that was more interpretable within the context of the sensitivity functions defined above.  Moreover, precise limits on each parameter were less of a concern for the current analysis as our primary goal was to demonstrate differences in fit as a function of sensitivity and identifiability.  The optimal values for each parameter were identified based on convergence of the model results as a minimization of the \ac{rmse} between the observed and modelled data.

@@

Calibration of parameter values for each of the seven selection heuristics (\cref{tab:heurist1,tab:heurist2,tab:heurist3})  produced model output that varied in goodness of fit and computational effort required to identify optimal values (\cref{tab:calib}).  As expected, the number of required iterations of the optimization algorithm to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only \Sexpr{toeng(temp$its)[[1]]} iterations were required to optimize the single parameter selected for the temperature category, whereas \Sexpr{top$its} iterations were required for the \Sexpr{toeng(top$nprm)[[1]]} parameters selected independent of category.  No association was observed between collinearity ($\gamma$) and number of iterations or calibration.  Of the seven parameter subsets, selection independent of parameter category and equally within each category produced the largest reduction in \ac{rmse} values from the default output (starting \ac{rmse} \Sexpr{strt} mmol O$_2$ m$^{-3}$, reduction to \Sexpr{scinot(top$rmse)}, \Sexpr{scinot(cattop$rmse)}, respectively), whereas the optimal values for parameters in the Optics category had the smallest reduction (\Sexpr{scinot(optics$rmse)}).  Lower \ac{rmse} was associated with higher average sensitivity of a parameter subset.  Parameters selected only in the optics category had the highest \ac{rmse} and lowest average L1 (\Sexpr{scinot(optics$avee)}, \Sexpr{toeng(optics$nprm)} parameters), whereas parameters selected independent of category had the lowest \ac{rmse} and highest average L1 (\Sexpr{scinot(top$avee)}, \Sexpr{toeng(top$nprm)} parameters, \cref{tab:calib}).

@@

State variables were most sensitive to phytoplankton and zooplankton parameters, particularly the maximum growth rates (\Sexpr{par_txt('umax_1')} for phytoplankton, \Sexpr{par_txt('Zumax_1')} for zooplankton), mortality coefficient for phytoplankton (\Sexpr{par_txt('mA_1')}), and the zooplankton half saturation coefficent for grazing (\Sexpr{par_txt('ZKa_1')}). An increase in the growth rate of primary producers can increase oxygen concentration through photosynthetic processes, although increased production of organic matter is balanced with respiration and bacterial decomposition that reduce \ac{do} in the water column.  Similarly, increases in zooplankton abundance with increased growth rates can reduce phytoplankton biomass through grazing, which is expected to further deplete pools of organic matter from algal sources.  Most variables were also sensitive to variation in the half-saturation grazing coefficient which moderates nutrient concentrations that support half the maximum grazing rate. Although the tradeoff between abundance, grazing, and decomposition is complex, the sensitivity of model state variables to parameters that directly control the abundance of primary producers is in agreement with empirical observations of factors that influence hypoxia dynamics on the \ac{lcs} \citep{Fahnenstiel95,Roelke00,Eldridge10}. The sensitivity of the model output to variation in other parameters that relate to physical and chemical properties of the system was of secondary importance.  That is, state variables were sensitive to changes in light and temperature parameters, although to a lesser extent than biological parameters.  As such, the differing sensitivities of state variables to parameters in each of the categories was not unexpected given ecological relationships that are well understood and described by the model.

A general conclusion from the identifiability analyses is that only limited subsets of parameters were identifiable within the constraints of local sensitivity analyses. These results support previous studies that have suggested similarly small subsets of parameters can be identified using traditional calibration schemes \citep[e.g.,][]{Wheater86,Ye97,Omlin01}.  In addition to \ac{cgem}, these conclusions have relevance for other biogeochemical models that include numerous parameters and structural equations to characterize processes in the model domain.  A general conclusion is that less complex models are potentially beneficial given that only a small subset of parameters is identifiable and that ecosystem processes may in fact be sufficiently characterized with few parameters \citep{Ye97}.  Conversely, others have argued that model complexity is not in itself a disadvantage when parsimony is not the only determinant of model structure \citep{Reichert97}. Over-parameterization can be useful if processes have importance that were not evaluated during model identification.  Single objective functions that maximize model fit with identifiable parameters may also provide an incomplete characterization of model worth, which has prompted the development of probability-based models of hypoxia that explicitly include uncertainty in model components \citep[e.g.][]{Obenour15}.  Our results demonstrated that approximately 75\% of the evaluated parameters had an effect on the eight state variables, whereas \ac{cgem} includes a total of 36 variables and multiple plankton groups, not all of which have immediate concern for understanding hypoxia. The redundancies identified with the sensitivity analyses are challenging only if the primary interest is, for example, \ac{do} dynamics.  Moreover, the proposed selection heuristics provide flexibility for choosing different parameters with the assumption that those chosen depend on the research or management question.
